---
title: "h5Seurat File Format Specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{h5Seurat File Format Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(hdf5r)
```

## Overall File Structure

### Required Attributes

There are three required top-level HDF5 attributes: `project`, `active.assay`, and `version`. Each of these must be a single [string](#string-representation). The `project` attribute corresponds to the project value of a `Seurat` object; the `active.assay` attribute is the name of the default assay and must be present in the [`assays` group](#assay-expression-data). The `version` corresponds to the version of Seurat that the h5Seurat file is based on.

### Top-Level Datasets and Groups

There are two required top-level HDF5 datasets: `cell.names` and `meta.data`. The `cell.names` dataset should be a one-dimensional [string](#string-representation) dataset, with a length equal to the number of cells present in the data. Cell names are not stored anywhere else in the h5Seurat file.

The `meta.data` dataset contains cell-level metadata. It should be a one-dimensional [HDF5 compound type](#data-frame-representation), with a length equal to the number of cells present in the data. Each component dataset in the compound corresponds to a column of metadata information (eg. number of UMIs per cell/`nFeature`) and should be a one-dimensional dataset corresponding to the nearest R type.

## Assay Expression Data

`Assay` objects are stored in the top-level group `assays`; each assay is stored as its own group within the `assays` group.

## Dimensional Reductions

## Nearest-Neighbor Graphs

## Spatial Image Data

## Cell-Level Metadata

## Command Logs

## Miscellaneous Information

## Tool-Specific Results

## Common Data Structures

### String Representation

All string values should be encoding as variable-length UTF-8 characters

```{r utf8, echo=FALSE}
H5T_STRING$new(size = Inf)$set_cset(cset = h5const$H5T_CSET_UTF8)
```

### Dense Matrix Representation

### Sparse Matrix Representation

### Factor Representation

```{r factor, echo=FALSE}
```

### Data Frame Representation

`data.frames` should be stored 

### Generic S4 Object Representation
